---
title: 'Assignment: Among Site (Beta) Diversity'
author: 'Evgeniya Polezhaeva; Z620: Quantitative Biodiversity, Indiana University'

output:
  html_document: default
  pdf_document: default
geometry: margin=2.54cm
---

## OVERVIEW

In this Assignment, we move beyond the investigation of within-site $\alpha$-diversity.
We will explore $\beta$-diversity, which is defined as the diversity that occurs among sites.
This requires that we examine the compositional similarity of assemblages that vary in space or time.

After completing this exercise you will know how to:

1. formally quantify $\beta$-diversity
2. visualize $\beta$-diversity with heatmaps, cluster analysis, and ordination
3. test hypotheses about $\beta$-diversity using multivariate statistics

## Directions:
1. Change "Student Name" on line 3 (above) with your name.
2. Complete as much of the exercise as possible during class; what you do not complete in class will need to be done on your own outside of class.
3. Use the Handout as a guide; it contains a more complete description of data sets along with the proper scripting needed to carry out the exercise.
4. Be sure to **answer the questions** in this exercise document; they also correspond to the Handout.
Space for your answer is provided in this document and indicated by the ">" character.
If you need a second paragraph be sure to start the first line with ">".
5. Before you leave the classroom, **push** this file to your GitHub repo.
6. When you are done wit the Assignment, **Knit** the text and code into a html file.
7. After Knitting, please submit the completed Assignment by creating a **pull request** via GitHub.
Your pull request should include this file *beta_assignment.Rmd* and the html output of `Knitr` (*beta_assignment.html*).

## 1) R SETUP

Typically, the first thing you will do in either an R script or an RMarkdown file is setup your environment. 
This includes things such as setting the working directory and loading any packages that you will need.

In the R code chunk below, provide the code to:  

1. clear your R environment,
2. print your current working directory,
3. set your working directory to your "*/Week3-Beta*" folder, and
4. load the `vegan` R package (be sure to install if needed).  

```{r}
rm(list=ls())
getwd()
setwd("D:/Jane/GitHub/QB2017_Polezhaeva/Week3-Beta")
require("vegan")
```

## 2) LOADING DATA
### Load dataset
In the R code chunk below, do the following:  

1. load the `doubs` dataset from the `ade4` package, and
2. explore the structure of the dataset.

```{r}
package.list <- c('vegan', 'ade4', 'viridis', 'gplots', 'BiodiversityR', 'indicspecies') 
for (package in package.list) {
  if (!require(package, character.only=T, quietly=T)) { 
    install.packages(package)
    library(package, character.only=T)
  } }

require("ade4")
data(doubs)
str(doubs, max.level = 1)
summary(doubs)
head(doubs$env)
head(doubs$fish)
```

***Question 1***:  Describe some of the attributes of the `doubs` dataset.  

a.  How many objects are in `doubs`?
b.  How many fish species are there in the `doubs` dataset?
c.  How many sites are in the `doubs` dataset?

> ***Answer 1a***:  4
> ***Answer 1b***:  27
> ***Answer 1c***:  30

### Visualizing the Doubs River Dataset

***Question 2***: Answer the following questions based on the spatial patterns of richness (i.e., $\alpha$-diversity) and Brown Trout (*Salmo trutta*) abundance in the Doubs River.  

a.  How does fish richness vary along the sampled reach of the Doubs River?
b.  How does Brown Trout (*Salmo trutta*) abundance vary along the sampled reach of the Doubs River?
c.  What do these patterns say about the limitations of using richness when examining patterns of biodiversity?

> ***Answer 2a***:  Species richness encreases unevenly from upstrem to downstrem region. It is highest in two regions: very downstream and inbetween of two turns of the river.
> ***Answer 2b***:  Brown trout is present only in two regions, where species richness is low. Brown Trout abundacne is in contrphase to species richness.
> ***Answer 2c***:  The concept of richness does not give us any information about particular species, but just total amount of all species without considering ratio between them. From sp. richness data we cannot conclude anything about particular species.

## 3) QUANTIFYING BETA-DIVERSITY

In the R code chunk below, do the following:  

1. write a function (`beta.w()`) to calculate Whittaker's $\beta$-diversity (i.e., $\beta_{w}$) that accepts a site-by-species matrix with optional arguments to specify pairwise turnover between two sites, and
2. use this function to analyze various aspects of $\beta$-diversity in the Doubs River.  

```{r}
beta.w <- function(site.by.species = "", sitenum1 = "", sitenum2 = "", pairwise = FALSE){
  #Only if pairwise is TRUE, do this:
  if (pairwise == TRUE){
    #As a check ,print an error if we do not provide needed arguments
    if (sitenum1=="" | sitenum2 ==""){
      print("Error:Please speify sites to compare")
      return(NA)}
    #Calculate pairwise betadiversity
    site1 = site.by.species[sitenum1,]
    site2 = site.by.species[sitenum2,]
    site1 = subset(site1, select = site1 > 0)
    site2 = subset(site2, select = site2 > 0)
    gamma = union(colnames(site1), colnames(site2))
    s = length(gamma)
    a.bar = mean(c(specnumber(site1), specnumber(site2)))
    b.w = round(s / a.bar - 1, 3)
    return(b.w)
  }
  else{
    SbyS.pa <- decostand(site.by.species, method = "pa")
    S <- ncol(SbyS.pa[, which(colSums(SbyS.pa) > 0)])
    a.bar <- mean(specnumber(SbyS.pa))
    b.w <- round(S / a.bar, 3)
    return(b.w)
  }
  }
beta.w(doubs$fish, sitenum1 = "1", sitenum2 = "2", pairwise =  T)
alpha <- specnumber(doubs$fish[,1]) + specnumber(doubs$fish[,2])
alpha
```

***Question 3***: Using your `beta.w()` function above, answer the following questions:

a.  Describe how local richness ($\alpha$) and turnover ($\beta$) contribute to regional ($\gamma$) fish diversity in the Doubs.
b.  Is the fish assemblage at site 1 more similar to the one at site 2 or site 10?
c.  Using your understanding of the equation $\beta_{w} = \gamma/\alpha$, how would your interpretation of $\beta$ change if we instead defined beta additively (i.e., $\beta = \gamma - \alpha$)?

> ***Answer 3a***:  regional fish diversity $\gamma$ is proportional to product of average local diversity ($\alpha$) and turnover between two sites increased by 1.
> ***Answer 3b***:  Site 1 is more similar to site 2 ( $\beta$ = 0.5) than 10 ( $\beta$ = 0.5)
> ***Answer 3c***:  If $\beta$ is defined additively it means that beta-diversity is number of species that are, in averege, present in regional species pool but not present in each site within the region. For example, if $\beta_{w} = 1$ it means that local site as diverse as regional pool and additive $\beta = 0$, but if additive $\beta = 1$ it means that local site has one less species than the region.

### The Resemblance Matrix
In order to quantify $\beta$-diversity for more than two samples, we need to introduce a new primary ecological data structure: the **Resemblance Matrix**. 

***Question 4***: How do incidence- and abundance-based metrics differ in their treatment of rare species?

> ***Answer 4***:  Since incidence matrix contains only presence-absence information, incidence-based metrics does not differentiate between rare and abundant species.

In the R code chunk below, do the following:  

1. make a new object, `fish`, containing the fish abundance data for the Doubs River, 
2. remove any sites where no fish were observed (i.e., rows with sum of zero),
3. construct a resemblance matrix based on Sørensen's Similarity ("fish.ds"), and
4. construct a resemblance matrix based on Bray-Curtis Distance ("fish.db").  

```{r}
fish <- doubs$fish
print(fish)
fish <- fish[-8,] #Removes row 8 which does not contain observations

require("vegan")
fish.ds <- vegdist(fish, method = "bray", binary = TRUE) #Construct Sorensen similarity resemblence matrix
fish.db <- vegdist(fish, method = "bray") # construct Bray-Curtis Distance resemlance matrix

fish.db <- vegdist(fish, method = "bray", diag = T)
fish.ds <- vegdist(fish, method = "bray", binary = TRUE, diag = T)
```

***Question 5***: Using the distance matrices from above, answer the following questions:

a. Does the resemblance matrix (`fish.db`) represent similarity or dissimilarity? What information in the resemblance matrix led you to arrive at your answer? 
b. Compare the resemblance matrices (`fish.db` or `fish.ds`) you just created. How does the choice of the Sørensen or Bray-Curtis distance influence your interpretation of site (dis)similarity?

> ***Answer 5a***:  The resemblance matrix fish.db represent dissimilarity (the diagonal elements (distance between site(i) and itself) are 0).
> ***Answer 5b***: Sorensen measure considers presence-absence data, while Bray-Curtis uses abundance data. Sorensen measure treats all species equally, unlike Bray-Curtis. For  more even samples this two measures would be closer than for uneven ones. 

## 4) VISUALIZING BETA-DIVERSITY

### A. Heatmaps

In the R code chunk below, do the following:  

1. define a color palette,
2. define the order of sites in the Doubs River, and
3. use the `levelplot()` function to create a heatmap of fish abundances in the Doubs River.

```{r}
#Define order of sites
order <- rev(attr(fish.db, "Labels"))

#Plot HeatMap
levelplot(as.matrix(fish.db)[, order], aspect = "iso", col.regions = inferno, xlab = "Doubs Site", ylab = "Doubs Site", scales = list(cex = 0.5), main = "Bray-Curtis Distance")

```

### B.  Cluster Analysis

In the R code chunk below, do the following:  

1. perform a cluster analysis using Ward's Clustering, and
2. plot your cluster analysis (use either `hclust` or `heatmap.2`). 

```{r} 
fish.ward <- hclust(fish.db, method = "ward.D2") #Perform cluster analysis

#Plot Cluster
par(mar = c(1, 5, 2, 2) + 0.1)
plot(fish.ward, main = "Doubs River Fish: Ward's Clustering", ylab = "Squared Bray-Curtis Distance")

gplots::heatmap.2(as.matrix(fish), distfun = function(x) vegdist(x, method = "bray"), hclustfun = function(x) hclust(x, method = "ward.D2"), col = viridis, trace = "none", density.info = "none")

```

***Question 6***: Based on cluster analyses and the introductory plots that we generated after loading the data, develop an ecological hypothesis for fish diversity the `doubs` data set? 

> ***Answer 6***: Based cluster analysis all sites can be divided into 4 groups. 1-14 : low diversity, but high abundance of present spesies, 15-19 : medium diversity, medium abundance; 20-30 (except 23-25) sites: high diversity and abundance of fish species, 23-25 sites: low diversity and abundance.  In general species diversity increases from upstream regions to downstream along with species abundance. There is only one region that does not follow this pattern (sites 23-25), that might be caused by environmental conditions.  

### C.  Ordination

### Principal Coordinates Analysis (PCoA)

In the R code chunk below, do the following:  

1. perform a Principal Coordinates Analysis to visualize beta-diversity
2. calculate the variation explained by the first three axes in your ordination
3. plot the PCoA ordination,
4. label the sites as points using the Doubs River site number, and
5. identify influential species and add species coordinates to PCoA plot.

```{r}
require("stats")
fish.pcoa <- cmdscale(fish.db, eig = TRUE, k = 3)

explainvar1 <- round(fish.pcoa$eig[1] / sum(fish.pcoa$eig), 3) * 100
explainvar2 <- round(fish.pcoa$eig[2] / sum(fish.pcoa$eig), 3) * 100
explainvar3 <- round(fish.pcoa$eig[3] / sum(fish.pcoa$eig), 3) * 100

sum.eig <- sum(explainvar1, explainvar2, explainvar3)
sum.eig

#define plot parameters
par(mar = c(5, 5 ,1, 2) + 0.1)

#Initiate plot
plot(fish.pcoa$points[ ,1], fish.pcoa$points[, 2], ylim = c(-0.2, 0.7), 
     xlab = paste("PCoA 1 (", explainvar1, "%)", sep = ""),
     ylab = paste("PCoA 2 (", explainvar2, "%)", sep = ""),
    pch = 16, cex = 2.0, type = "n", cex.lab = 1.5, cex.axis = 1.2, axes = FALSE)

#Add axes
axis(side = 1, labels = T, lwd.ticks = 2, cex.axes = 1.2, las = 1)
axis(side = 2, labels = T, lwd.ticks = 2, cex.axes = 1.2, las = 1)
abline(h = 0, v = 0, lty = 3)
box(lwd = 2)

#Add points and labels
points(fish.pcoa$points[,1], fish.pcoa$points[, 2], pch = 19, cex = 3, bg = "gray", col = "gray")
text(fish.pcoa$points[,1], fish.pcoa$points[,2], labels = row.names(fish.pcoa$points))

#Relative abundance of e sp at e site
fishREL <- fish
  for (i in 1:nrow(fish)){
    fishREL[i, ] = fish[i, ] / sum(fish[i, ])
  }
#Calc and add sp scores
fish.pcoa <- add.spec.scores(fish.pcoa, fishREL, method = "pcoa.scores")
text(fish.pcoa$cproj[, 1], fish.pcoa$cproj[, 2], labels = row.names(fish.pcoa$cproj), col = "black")
```

In the R code chunk below, do the following:  

1. identify influential species based on correlations along each PCoA axis (use a cutoff of 0.70), and
2. use a permutation test (999 permutations) to test the correlations of each species along each axis.

```{r}
spe.corr <- add.spec.scores(fish.pcoa, fishREL, method = "cor.scores")$cproj
corrcut <- 0.7 #cutoff
imp.spp <- spe.corr[abs(spe.corr[,1]) >= corrcut | abs(spe.corr[,2]) >= corrcut, ]

fit <- envfit(fish.pcoa, fishREL, perm = 999) #Permutation Test for Species Abundances Across Axes
imp.spp
fit
```

***Question 7***: Address the following questions about the ordination results of the `doubs` data set:

a.  Describe the grouping of sites in the Doubs River based on fish community composition.
b.  Generate a hypothesis about which fish species are potential indicators of river quality.

> ***Answer 7a***:  There are 4 groups of sites base on fish community composition: sites 1-14, sites 15-19, sites 23-25, sites 20-30 (ex.23-25). The first 3 groups differ along only 1 axes. The group 4 stand far away along the second axes.
> ***Answer 7b***: High qulity sites may be identified by presence of species Sart, Phph. an Sites 23-25 belong to region of low quality and the species Alal the determins its position on PCoA plot.


## 5) HYPOTHESIS TESTING

### A.  Multivariate Procedures for Categorical Designs
Earlier work done in the Doubs River suggested that the river has four distinct regions of habitat quality: the first region (sites 1-14) of "high quality"; the second (sites 15 - 19) and fourth (sites 26 - 30) of "moderate quality"; and the third  (sites 20 - 25) of "low quality".

In the code chunk below, test the hypothesis that fish community composition varies with river quality.

1. create a factor vector that categorizes habitat quality in the Doubs River,
2. use the multivariate analyses for categorical predictors to describe how fish community structure relates to habitat quality.

```{r}
#Create "Fator" vector
quality <- c(rep("HQ", 13), rep("MQ", 5), rep("LQ", 6), rep("MQ", 5))

#Run PERMANOVA with adonis function
adonis(fish ~ quality, method = "bray", permutations = 999)

#significant species for id habitat quality
indval <- multipatt(fish, cluster = quality, func = "IndVal.g", control = how(nperm=999))
summary(indval)

#habitat preferences of species
fish.rel <- decostand(fish, method = "total")
phi <- multipatt(fish.rel, cluster = quality, func = "r.g", control=how(nperm=999))
summary(phi)
```

***Question 8***: Based on the PERMANOVA, IndVal, and phi coefficient analyses, what did you learn about the relationship between habitat quality and the fish species composition?

> ***Answer 8***: Data of IngVal and phi coefficient analysis supported conclusions drawn from PERMANOVA. There are certain species that more likely to be present in certain groups of habitats. Phph and Satr in  HQ sites, Alal and  Ruru  in LQ sites. List of charactieristic species  of  Medium quality habitat differs depending on type o fanalysis used.


### B.  Multivariate Procedures for Continuous Designs

### i.  Mantel Test
In the R code chunk below, do the following:   

1. create distance matrices for both fish communities and environmental factors, and
2. use a Mantel test to determine if these matrices are correlated, and test the hypothesis that fish assemblages are correlated with stream environmental variables.

```{r}
#defining matrices
fish.dist <- vegdist(doubs$fish[-8,], methos = "bray")
env.dist <- vegdist(scale(doubs$env[-8,]), method = "euclid")

#Mantel test
mantel(fish.dist, env.dist)

```

***Question 9***: What do the results from our Mantel test suggest about fish diversity and stream environmental conditions? How does this relate to your hypothesis about stream quality influencing fish communities?

> ***Answer 9***:  Fish diversity and Stream environment conditions positevely correlate with each other (r= 0.604 > 0) with very low p-value that is consistent with the hypothesis about stream quality (environmental conditions) influencing fish communities.

### ii.  Constrained Ordination
In the R code chunk below, do the following:  

1. create an environmental matrix of the water chemistry data included in the `doubs` dataset using forward and reverse selection of variables,
2. conduct a redundancy analysis on the fish assemblages of the Doubs River,
3. use a permutation test to determine the significance of the constrained analysis,
4. use a permutation test to determine the correlation of each environmental factor on the constrained axes,
5. calculate the explained variation on the first and second constrained axes,
6. plot the constrained ordination results including labeled points for each site, and
7. add vectors that demonstrate the influence of each environmental factor the constrained ordination.   

```{r}
#1.
env.chem <- as.matrix(doubs$env[-8, 5:11])
#perform dbRDA
doubs.dbrda <- dbrda(fish.db ~ ., as.data.frame(env.chem))
ordiplot(doubs.dbrda)

psych::corr.test(env.chem)

#Model only the intercept
doubs.dbrda.mod0 <- dbrda(fish.db ~ 1, as.data.frame(env.chem))
ordiplot(doubs.dbrda.mod0)

#Model the full model with all explanatory variables
doubs.dbrda.mod1 <- dbrda(fish.db ~ ., as.data.frame(env.chem))
doubs.dbrda <- ordiR2step(doubs.dbrda.mod0, doubs.dbrda.mod1, perm.max = 200)

doubs.dbrda$call
doubs.dbrda$anova
ordiplot(doubs.dbrda)

#Perm test to evaluate significance
permutest(doubs.dbrda, permutations = 999)
envfit(doubs.dbrda, env.chem[, c(4,6,7)], perm = 999)

dbrda.explainvar1 <- round(doubs.dbrda$CCA$eig[1] / sum(c(doubs.dbrda$CCA$eig, doubs.dbrda$CA$eig)), 3) * 100

dbrda.explainvar2 <- round(doubs.dbrda$CCA$eig[2] / sum(c(doubs.dbrda$CCA$eig, doubs.dbrda$CA$eig)), 3) * 100

par(mar = c(5,5,4,4) + 0.1)

plot(scores(doubs.dbrda, display="wa"), xlim=c(-1.3, 1.1), ylim=c(-1.1, 2.7), xlab=paste("dbRDA 1 (", dbrda.explainvar1, "%)", sep=""), ylab=paste("dbRDA 2 (", dbrda.explainvar2, "%)", sep=""), pch=16, cex=2.0, type="n", cex.lab=1.5, cex.axis=1.2, axes=F)

#Add axes
axis(side=1, labels = T, lwd.ticks=2, ces.axis=1.2, las=1)
axis(side=2, labels = T, lwd.ticks=2, ces.axis=1.2, las=1)
abline(h = 0, v = 0, lty = 3)
box(lwd = 2)
points(scores(doubs.dbrda, display = "wa"), pch = 19, cex = 3, bg = "gray", col = "gray") 
text(scores(doubs.dbrda, display = "wa"), labels = row.names(scores(doubs.dbrda, display = "wa")))
# Add Environmental Vectors 
vectors <- scores(doubs.dbrda, display = "bp")

arrows(0, 0, vectors[,1], vectors[, 2], lwd = 2, lty = 1, length = 0.2, col = "red")

text(vectors[,1], vectors[, 2], pos = 3, labels = row.names(vectors)) 

axis(side = 3, lwd.ticks=2, cex.axis=1.2, las = 1, col = "red", lwd = 2.2, at = pretty(range(vectors[, 1])) * 2, labels = pretty(range(vectors[, 1]))) 

axis(side = 4, lwd.ticks=2, cex.axis=1.2, las = 1, col = "red", lwd = 2.2, at = pretty(range(vectors[, 2])) * 2, labels = pretty(range(vectors[, 2])))


```

***Question 10***: Based on the constrained ordination, what are the environmental variables (or groups of correlated variables) that seem to be contributing to variation in fish community structure?

> ***Answer 10***: The following parameters contribute the most to variation in community structure: oxy, bdo, nit. Parameter bdo correlates with pho and amm, parameter nit - with pho and amm too, and parameter oxy strongly negatively correlate with pho and bdo. 

### iii. Variation Partitioning
In the code chunk below,   

1. Create a matrix model of the selected environmental variables,
2. Create a matrix model of the selected PCNM axes,
3. Perform constrained and partial constrained ordinations using the spatial and environmental models you just created,
4. Test the significance of each of your constrained ordinations using permutation tests,
5. Partition the variation among sites into the relative importance of space, environment, spatially structured environment, and residuals,
6. Plot the variation partitioning output to visualize it.  
```{r}
doubs.dbrda$anova
env.mod <- model.matrix(~ oxy + bdo + nit, as.data.frame(env.chem))[,-1]

#Weight each site by its relative abundance 
rs <- rowSums(fish)/sum(fish)
#Perform PCNM
doubs.pcnmw <- pcnm(dist(doubs$xy[-8,]), w = rs, dist.ret = T)
# only eigenvectors associated with the positive eigenvalues are meaningful
doubs.pcnmw$values > 0 
doubs.space <- as.data.frame(scores(doubs.pcnmw))
doubs.pcnm.mod0 <- dbrda(fish.db ~ 1, doubs.space)
doubs.pcnm.mod1 <- dbrda(fish.db ~ ., doubs.space)
step.pcnm <- ordiR2step(doubs.pcnm.mod0, doubs.pcnm.mod1, perm.max = 200)
plot(step.pcnm)
step.pcnm$anova
space.mod <- model.matrix(~ PCNM2 + PCNM3 + PCNM5 + PCNM1 + PCNM13 + PCNM16 + PCNM6, doubs.space)[,-1] 


doubs.total.env <- dbrda(fish.db ~ env.mod)
doubs.total.space <- dbrda(fish.db ~ space.mod)
doubs.env.cond.space <- dbrda(fish.db ~ env.mod + Condition(space.mod))
doubs.space.cond.env <- dbrda(fish.db ~ space.mod + Condition(env.mod))

permutest(doubs.env.cond.space, permutations = 999)
permutest(doubs.space.cond.env, permutations = 999)
permutest(doubs.total.env, permutations = 999)
permutest(doubs.total.space, permutations = 999) 
doubs.varpart <- varpart(fish.db, env.mod, space.mod)
doubs.varpart

plot.new()
par(mar = c(2,2,2,2))
plot(doubs.varpart)
text(1, 0.25,"Space")
text(0, 0.25, "Env")
mtext("Variation Partitioning of\nDoubs Fish Diversity", side = 3, line = -1)


```

***Question 11***: Interpret the variation partitioning results.   

> ***Answer 11***: Based on the variation partitioning analysis 13% of Fish Diversity can be explained by only environmental component, 24% - by space alone, 37% - by both space and anvironment, and 27% of variation of fish diversity cannot be explained by environment or space.

## SYNTHESIS
Load the dataset you are using for your project. Perform an ordination to visualize your dataset.  Using this ordination, develop some hypotheses relevant to $\beta$-diversity and identify the appropriate tools you would use to test them.